---
title: "SPINK"
author: "Fan Yue"
date: "2026/04/02"
---

## 1. Overview
SPINK is an R package for inferring **enhancer-gene regulatory interactions** from spatially resolved multi-ome data. Built on a **hierarchical linear spatial model (HLSM)** with thin-plate splines, SPINK explicitly models spatial dependence to deliver well-calibrated p-values and superior statistical power compared to existing methods.

## 2. Installation
TDEseq is implemented as an R package, which can be installed from GitHub.
```
library(devtools)
install_github("fanyue322/SPINK")
```
## 3. Usage
TDEseq starts with spatial multi-ome data with RNA assay and ATAC assay . 

### Application on human hippocampus data
We used a human hippocampus data as a main example which contains expression measurement of 29,293 genes and 42,944 peaks on 2500 spots. This dataset can be downloaded from their original study GSE205055. We also saved the processed data that we used in our examples in RData format, which can be downloaded from [here](https://zenodo.org/record/8276740/files/scRNA-seq_liver_embryo.rda). A detailed protocol for constructing a Spatial multiome object in the Seurat format is available [here](https://stuartlab.org/signac/articles/pbmc_multiomic)
```{r}
#| eval: false
library(Seurat)
library(Signac)
object<-readRDS("HumanHippocampus_spatial_RNA_ATAC.rds")
```

#### Processing 
It is encouraged using fragment counts for enhancer-gene link analysis. Users can calling fragment counts with Signac. To use the peak calling functionality in Signac you will first need to install MACS2. This can be done using pip or conda, see the detail from [here](https://github.com/macs3-project/MACS). Using "which" to check the path `which macs2`.
```{r}
#| eval: false
DefaultAssay(object)='ATAC'
peaks <- CallPeaks(object, macs2.path = "~/anaconda3/bin/macs2")
peaks <- keepStandardChromosomes(peaks, pruning.mode = "coarse")
peaks <- subsetByOverlaps(x = peaks, ranges = blacklist_hg38_unified, invert = TRUE)
macs2_counts <- FeatureMatrix(
  fragments = Fragments(object),
  features = peaks,
  cells = colnames(object)
)

fragments="./GSM6206884_HumanBrain_50um_fragments.tsv.gz"
library(EnsDb.Hsapiens.v86)
library(AnnotationHub)
library(ensembldb)
library(AnnotationForge)
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotations) <- paste0('chr', seqlevels(annotations))
genome(annotations) <- "hg38"
Annotation(object) <- annotations


object[["peaks"]] <- CreateChromatinAssay(
  counts = macs2_counts,
  fragments = fragments,
  annotation = annotations
)
```

# Under Construction

